From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions

### 2008-08-31 thesis/paper.tex
Final versions of sec 1 and 2

As access to vast amounts of information becomes more and more ubiquitous, and
the information accessible to humans grows more diverse every day, users
struggle to find what they need among the vast amounts of data available to
them. Current search technologies aim to provide easy access to a constantly
changing pool of knowledge, yet they are hardly beyond the stage of looking for
exact text matches. Machine processing still fails to \emph{understand} the
information it has to deal with and sees data as nothing more than a set of
strings. But often a user's demands do not ask for a piece of text, but for an
abstract concept, or a relation. Knowing that natural language is most likely
infinite, such relations are virtually endless in number, and  one can only
conclude that trying to enumerate all relations between the bits of  human
knowledge in gargantuan databases is bound to be a never-ending task.

Growing efforts have thus been made to provide a technology that is (going to
be) able to extract logical formulae from sentences of ordinary language and
establish relations between objects of their logical representation. In addition,
providing a means to compute with said representations will eventually enable
the machine-processing of an abstract concept of meaning. In doing so, the
ultimate goal may be to enable a man-made system to \emph{understand} what
humans communicate. Then, these representations of knowledge could be queried, and
infer from their knowledge, what an optimal answer to a given query might be --
maybe even inventing an answer to a question never posed before along the way.

@@

The present work will aim to provide the \curt{} system of
\cite{blackburnbos:cl1} with the mechanics of dealing with more than one
partition of its logical domain, viz. the ability to handle more than one
\term{possible} or \term{accessible world}. As a ground for the development of this
work, the treatment of \term{embedded questions} was chosen as one that
naturally demands a possibility to account for more than one division in a
system's \emph{understanding} or representation of a universe of objects and
relations.

As the scope of a Bachelor's thesis is limited by its nature, not all of the
theory and reasoning can be understood without background knowledge in some
rather divergent domains. In particular, a basic familiarity with traditional
\term{natural language semantics} and \term{extensional} as well as
\term{intensional logic} or \term{modal} frameworks will be assumed.
Furthermore, a basic acquaintance with \term{eroteric logic} will be helpful,
but not strictly necessary in understanding some of the reasoning in section
\ref{sec:theory}.  Some knowledge of the \prol{} programming language and a good
overview of the architecture of the \curt{} system is imperative in order to
understand the implementation itself.

The first section will give a conceptual overview of the technologies to date
and the particular software, \curt, that will serve as a basis for extension.
The next section concerns itself with a more detailed analysis of the theoretic
assumptions of this implementation and justifies some choices and restrictions
that had to be made due to the limited scope of this analysis. While the
third section will display and document the actual implementation on top of the
\curt{} system, the last section will discuss the results of this work and also
give a prospect on possible further work on the subject.

@@

\begin{itemize}
  \item Text in a bold face sans serif font marks new terminology and important
  concepts: \term{embedded interrogatives}, \term{backtracking}
  \item Proper names, names of technologies, and abbreviations are typeset in a
  regular sans serif font: \acurt, \prol, \pn{Ty2}
  \item Code is written in a monospace font, resembling the look of a font class most
  programmers are used to:\\
  \code{lam(P, lam(Q, all(X, imp(app(P,X),app(Q,X)))))}.
  \item Logical formulae are typeset in a different serif font from the rest of
  the text:\\
    Assume $k = (\lambda x.\neg xx)$. Then $ kk = (\lambda x .\neg xx)k = \neg kk$
  \item Quotations  are typeset on a separate paragraph, using a sans serif font:
  \quotesf{This answers the question!}
\end{itemize}

\section{Concepts}

@@

This section delineates the ground the software presented in this thesis is
based on. \ref{sec:curt} introduces the base system of the implementation,
\curt{},
and \ref{sec:comparison} relates it to other similar systems in existence.
\ref{sec:extension} then goes on to explain where the \curt{} system lacks in
features and thus exhibits potential for extension.

@@

Besides parsing and translating utterances into logic, \curt{} is able to perform
basic checks for \emph{consistency} and \emph{informativity} on the resulting
formulae -- within the bounds of the tools it uses.
It can thus detect incoherent discourse and would reject a sequence of
sentences such as \emph{``Mary likes every man. Peter is a man. Mary does not
like Peter''}, claiming it found a contradiction in its input. The discourse is
then said to be \term{inconsistent} with the new input.

@@

The final chapter of \cite{blackburnbos:cl1} introduces the implementation
incrementally, presenting new versions of \curt{} that add to the software's
feature set one at a time. Support for inferencing via external theorem provers
and model builders is added right after the first, rather limited implementations.
Finally an ontology and a straight-forward way of parsing and answering direct
questions for noun phrases conclude its development. The theory and
implementation of this paper are meant as a replacement and enhancement for the
latter feature. Their introduction will also happen in an incremental manner and new features
are introduced one at a time during section \ref{sec:implementation}.

@@

The purpose of the former is in most cases twofold, although present
\abbr{QA} systems vary in the functionality they offer
and results they actually deliver. One goal is to enable their users to
query for information without the need for a special query language or at least
using a query language that would be easy and natural to use even for the
uninitiated user. This way the user can pose direct questions to a system
instead of having to search for a keyword to occur in the plain text of a
document, like they would have to do with traditional information
retrieval systems. The second feature provides \abbr{QA} systems with another important
distinction, viz. the ability to utilize ontologies in order to relate query
terms akin, but not equal to words in target documents and thus deliver
satisfactory results even for fuzzy queries.

@@

\cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
corpus-based \abbr{QA} techniques with deep semantic parsing based on
\term{Discourse Representation Theory}. Unfortunately, there seems
to be no working demo to date. The homepage of the ongoing efforts of Johan Bos
and his team can be found at the following address:
\href{http://svn.ask.it.usyd.edu.au/trac/candc/}{\url{http://svn.ask.it.usyd.edu.au/trac/candc/}}.

\cite{tenCate} present a tableaux method for resolving questions and also
propose an algorithm in \prol{} as an implementation. Their approach is a more
theoretical one, and they do not provide any particular implementation of their
work. Although their methods seems promising, to say the least, it was chosen to
avoid the implementation of a specialized inferencing system. \cite{bosgabdsil}
mention the combinatorial explosion that lurks behind the treatment of questions
and this is a considerable threat to any implementation of deep semantic
\abbr{QA} systems. By using external inferencing tools, an implementation of
the semantics of questions can focus on more essential issues than the
resolution of formulae. % too offending. fixme

Discourse Representation Theory is also used in \cite{bosgabdsil}, who present
an approach to question answering quite similar to what is presented in this
paper. In contrast to \cite{tenCate}, who use their own internal inferencing
method to resolve questions, Bos and Gabdsil also use external inferencing very
similar to the \curt{} system. Their results will be discussed later on, and
explicitly compared to the present results.

An attempt to bring inferencing to the world of \abbr{QA} systems can be found
in \cite{webber}.

Interestingly, earlier attempts at providing a software capable of answering
questions in natural language were mostly coupled with the usage of
semantic analysis instead of statistical methods.
\cite{chat} are among  the first to attempt creating a question answering system
with reasonable coverage. Their effort has been focused on the geographical
domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
\prol. The lack of hardware resources of that time forced the system to be
written with efficiency in mind. Moreover, it had to maintain a statical
database of facts which was hand-crafted. As such, it falls more into the
category of question answering systems than dialogue systems.

Today it is clear that statistical question answering systems make inroads to
the search market, and not many doubt that they are bound to eventually replace
current document/term-based search engines. Even though \abbr{QA} systems also
operate on the domain of answering questions, their approach to resolving them
differs in most aspects from that of dialogue systems. The latter have a much
higher obligation to correctness, since their results are typically not subject
to peer review and serve an entirely different purpose.

@@

\curt{} is a prototype dialogue system and as such it is based on an entirely different set of
premises than \abbr{QA} systems. Moreover, it differs from \pn{Godot} in some
major points, the most distinctive of which is that \curt{} only exists inside a
regular computer and has no access to sensory input. Also, \curt's
representation of knowledge differs from that of \pn{Godot} or the \pn{Pronto QA
System} in that it does not use discourse representation theory (yet). \acurt{}
addresses some of its shortcomings and lays ground to the implementation of a
broader feature set for it.

@@

At the end of \cite{blackburnbos:cl1}, \curt{} is able to cope with a small
grammar and provides some features for interaction in natural
language. It does only understand basic extensional transitive and intransitive
verbs and its support for question answering remains limited.
Anaphora resolution still remains an open issue.

@@

For this to work, the internal storage of \curt{} will have to be restructured
significantly in order to deal with more than one possible world or situation.
Its grammar will have to be enhanced so that it can handle \wh-embedding verbs
like `know' and `believe' and its inferencing mechanisms are going to be
employed in a different manner, to encompass a wider range of functionality.

@@

The semantics of questions has traditionally been a very sophisticated field of
linguistics,  philosophy and logics, with many different approaches arguing for
and against each other. This section will not try to give an exhausting overview
of the field of question semantics since this would be a task of considerable
size and by far beyond the scope of this thesis\footnote{For a good summary of
some of the literature, particularly the issue of whether questions are a
concern of pragmatics or semantics, consider \cite[section 2 and 3]{gs:q}.}.
Instead, it was chosen to roughly follow the work by Groenendijk and Stokhof,
who present one of the best-established and recognized theories in the field,
and also lending from other places where deemed appropriate or practical. As
such, the theoretical framework of \acurt{} is not going to be an exact
realization of any particular approach to question semantics, but rather an
accumulation of interesting concepts from different sources. The main sources,
however, are \cite{gs:q} and \cite{gs:sqpa}.

@@

\begin{termin}
  An \term{interrogative sentence}, or simply \term{interrogative}, is a
  type of sentence in natural language
  distinguished by certain characteristics
  \footnote{To quote \cite{gs:q}: \quotesf{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.},
  whereas an \term{interrogative act} refers to the utterance of an
  \term{interrogative}. Finally, a
  \term{question} will denote the semantic entity uttered by the
  \term{interrogative act}, as it will be introduced in section
  \ref{sec:protoq}
\end{termin}

@@

Example \ref{ex:yesno} is considered an \term{alternative question} or
\term{yes/no-interrogative} since only \emph{yes} and \emph{no} are possible
answers to this interrogative. \ref{ex:const} on the other hand is considered a
\term{\wh-question} or \term{constituent interrogative} since one of the
sentences' nominal constituents is replaced by a \wh-pronoun, prompting an
enumeration of possible fillers for the resulting gap as an answer.

The following abbreviations will be used: 
\abbr{Ty2} stands for two-sorted type logic\footnote{See \cite{gallin:ty2} for a
discussion and \cite{gs:sqpa} for its usage in describing the semantics of
questions}, \abbr{IL} stands for intensional logic\footnote{As used by
\cite{ptq}. See also \cite{gamut:2} for an excellent introduction to \abbr{IL}.},
\abbr{FOL}  is used as an abbreviation for first order logic. As an alternative
term for \term{possible world}, sometimes the expression \term{index} is used.

@@

Some research on the logic of questions (sometimes also referred to as eroteric
logic) suggests that the treatment of interrogatives demands its own formalism.
\acurt{} employs a formalism that departs from both \abbr{IL} and \abbr{Ty2} in
some ways. Most notably, the notion of a \term{model} is changed as follows, in
order to accommodate the question operator introduced further on, which demands
a set of possible worlds to be available.

@@

>>>>

As access to vast amounts of information becomes increasingly ubiquitous, and
the information accessible to humans grows more diverse every day, users
struggle to find what they need among the vast amounts of data available.
Current search technologies aim to address this problem by providing quick and
easy access to a constantly changing pool of knowledge, yet they are hardly
beyond the stage of looking for exact text matches. Machine processing still
fails to \emph{understand} the information it has to deal with and often sees data as
nothing more than a set of strings. But what if a user's demands do not ask for a
piece of text, but for an abstract concept, or a relation? Knowing that natural
language is most likely infinite, such relations are virtually endless in
number, and  one can only conclude that trying to enumerate all relations
between the bits of  human knowledge in gargantuan databases is bound to be a
never-ending task.

Growing efforts have thus been made to provide a technology that is (going to
be) able to extract logical formulae from sentences of ordinary language and
establish relations between objects of their logical representation. In
addition, providing a means to compute with said representations will eventually
enable the machine-processing of an abstract concept of meaning. In doing so,
the ultimate goal may be to enable a man-made system to \emph{understand} what
humans communicate.  These systems could then be queried and, given a capability
of performing logical inference on their domains of knowledge, they could
compute what an optimal answer to any given query might be -- perhaps even
by inventing an answer to a question never posed before.

@@

The present work aims to provide the \curt{} system of
\cite{blackburnbos:cl1} with the mechanics of dealing with \term{embedded
questions}, which enables it to reason not only about its own knowledge, but
also about the knowledge of individuals in its target domain.
For this, the system will have to enhanced with the capability of handling more
than one partition of its logical domain, viz. the ability to handle several
\term{possible} or \term{accessible worlds}.

As the scope of a Bachelor's thesis is limited by its nature, not all of the
theory and reasoning can be explained and background knowledge in some rather
divergent domains is assumed. In particular, a basic familiarity with
traditional \term{natural language semantics} and \term{extensional} as well as
\term{intensional logic} or \term{modal} frameworks will be assumed.
Furthermore, a basic acquaintance with theories of the logic of questions will be helpful,
but not strictly necessary in understanding some of the reasoning in section
\ref{sec:theory}.  Some knowledge of the \prol{} programming language and a good
overview of the architecture of the \curt{} system is imperative in order to
understand the implementation itself.

Section \ref{sec:concepts}  gives a conceptual overview of the technologies to
date and the particular software, \curt, that will serve as a basis for
extension.  The next section concerns itself with a more detailed analysis of
the theoretic assumptions of this implementation and justifies some choices and
restrictions that had to be made due to the limited scope of this analysis.
While the section \ref{sec:implementation} displays and documents the actual
implementation on top of the \curt{} system, the last section discusses the
results of the previous sections and also give a prospect on possible further work on the
subject. Finally, the appendix holds the sources code and some example
computations and models that show the basic functionality of \curt.

@@

\begin{itemize}
  \item Text in a bold face sans serif font marks new terminology and important
  concepts: \term{embedded interrogatives}, \term{backtracking}
  \item Proper names, names of technologies, and abbreviations are typeset in a
  regular sans serif font: \acurt, \prol, \pn{Ty2}
  \item Code is written in a monospace font, resembling the look of a font class most
  programmers are used to:
  \begin{lstlisting}
    lam(P, lam(Q, all(X, imp(app(P,X),app(Q,X)))))
  \end{lstlisting}
  \item Logical formulae are typeset in a different serif font from the rest of
  the text:\\
    Assume $k = (\lambda x.\neg xx)$. Then $ kk = (\lambda x .\neg xx)k = \neg kk$
  \item Quotations  are typeset on a separate paragraph, using a sans serif font:
  \quotesf{This answers the question!}
\end{itemize}

\section{Concepts}\label{sec:concepts}

@@

This section delineates the ground the software presented in this thesis is
based on while putting it into this rapidly evolving context. \ref{sec:curt}
introduces the base system of the implementation, \curt{}, and
\ref{sec:comparison} relates it to other similar systems in existence.
\ref{sec:extension} then goes on to explain where the \curt{} system lacks in
features and thus exhibits potential for extension.

@@

Besides parsing and translating utterances into logic, \curt{} is able to perform
basic checks for \emph{consistency} and \emph{informativity} on the resulting
formulae -- within the bounds of the tools it uses\footnote{\curt{} makes use
of (among others) the \pn{Otter} theorem prover and the \pn{mace2} model
builder. During the process of the implementation of section
\ref{sec:implementation}, it was ported to also make use of the \pn{Prover9}
theorem prover.}.
It can thus detect incoherent discourse and would reject a sequence of
sentences such as \emph{``Mary likes every man. Peter is a man. Mary does not
like Peter''}, claiming it found a contradiction in its input. The discourse is
then said to be \term{inconsistent} with the new input.

@@

The final chapter of \cite{blackburnbos:cl1} introduces the implementation
incrementally, presenting new versions of \curt{} that add to the software's
feature set one at a time. Support for inferencing via external theorem provers
and model builders is added right after the first, rather limited implementations.
Finally an ontology and a straight-forward way of parsing and answering direct
questions for noun phrases conclude its development. The theory and
implementation of this paper are meant as a replacement and enhancement for the
latter feature. 
%Their introduction will also happen in an incremental manner and new features
%are introduced one at a time during section \ref{sec:implementation}.

@@

The purpose of the former is in most cases twofold, although present \abbr{QA}
systems vary in the functionality they offer and results they actually deliver.
One goal is to enable their users to query for more complex information without
the need for a special query language or at least using a query language that
would be easy and natural to use even for the uninitiated user. The second
feature provides \abbr{QA} systems with another important distinction, viz. the
ability to utilize ontologies in order to relate query terms akin, but not equal
to words in target documents and thus deliver satisfactory results even for
fuzzy queries. This way the user can pose direct questions to a system instead
of having to search for a keyword to occur in the plain text of a document, like
they would have to do with traditional information retrieval systems.

@@

\cite{webber}, argue that \abbr{QA} systems would undoubtedly benefit from
introducing inferencing to their capability set. They analyse some of the
typical \abbr{QA} tasks and give examples to show that \emph{reasoning},
i.e. logical inferencing on the unstructured text that underlies modern
\abbr{QA} systems would enhance their functionality.

Interestingly, early attempts at providing a software capable of answering
questions in natural language were mostly coupled with the usage of semantic
analysis rather than statistical methods.  \cite{chat} are among  the first to
attempt creating a question answering system with reasonable coverage. Their
effort has been focused on the geographical domain, and their implementation,
dubbed \pn{CHAT-80}, is written entirely in \prol. The lack of hardware
resources of that time forced the system to be written with efficiency in mind.
Moreover, it had to maintain a statical database of facts which was
hand-crafted. As such, it falls more into the category of question answering
systems than dialogue systems. Within its domain, \pn{CHAT-80} can deliver quite
impressive results for different types of questions, including questions about
the nature of relations between objects in its database.  This way, the system
can answer a user's question as \example{Which country is borders at least two
oceans?} without having a list of countries that border two oceans at its
disposal.

\cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
corpus-based \abbr{QA} techniques with deep semantic parsing, utilizing
\term{Discourse Representation Theory} as a data structure. Unfortunately, there
seems to be no working demo to date. The homepage of the ongoing efforts of
Johan Bos and his team can be found at the following address:
\href{http://svn.ask.it.usyd.edu.au/trac/candc/}{\url{http://svn.ask.it.usyd.edu.au/trac/candc/}}.

Discourse Representation Theory is also used in \cite{bosgabdsil}, who present
an approach to question answering to a certain extend similar to what is pursued
in this paper. They choose to pursue the theoretical background compiled by
\cite{gs:sqpa} and several later publications, including \cite{gs:q}. Moreover,
they also use external first-order inferencing tools to perform computations on
the consistency and informativity of the system's DRS models, which for this
purpose are first translated to first order logic.  Their results will be
discussed later on, and explicitly compared to the present results.

\cite{tenCate} present a tableaux method for resolving questions via inferenceand also
propose an algorithm in \prol{} as an implementation. Although their methods
seems promising, it was chosen to
the present approach will avoid the implementation of a specialized inferencing
system and rely on external inferencing tools.

Today it is clear that question answering systems make inroads to the search
market, and not many doubt that they are bound to eventually replace current
document/term-based search engines. Dialogue systems, are still not capable of
convincing a larger user base. Among other things, their representation of
knowledge and capability to interact with stored information is still considered
to be lacking.  Even though \abbr{QA} systems also operate on the domain of
answering questions, their approach to resolving them differs from that of
dialogue systems. The latter have a much higher obligation to correctness, since
their results are typically not subject to peer review and serve an entirely
different purpose.

@@

\curt{} is a prototype teaching system, designed to showcase the possibilities
of a full-fledged dialogue system. It differs from \pn{Godot} in some major
points, the most distinctive of which is that \curt{} only exists inside a
regular computer and has no access to sensory input. Also, \curt's
representation of knowledge differs from that of \pn{Godot} or the \pn{Pronto QA
System} in that it does not use discourse representation theory (yet). \acurt{}
addresses some of its shortcomings and lays ground to the implementation of a
broader feature set for it.

@@

At the end of \cite{blackburnbos:cl1}, \curt{} is able to cope with a small
grammar and provides some features for interaction in natural
language. It does only understand basic extensional transitive and intransitive
verbs and its support for question answering is only rudimentary.
Anaphora resolution still remains an open issue.

@@

For this to work, the internal storage of \curt{} had to be restructured
significantly in order to deal with more than one possible world or situation.
Its grammar had to be enhanced so that it can handle \wh-embedding verbs
like \example{know} and \example{believe} and its inferencing functionality had
to be ported to the new system.

The next two sections explain, in order, the theoretical background and
practical operation of \acurt{}.

@@

The semantics of questions have traditionally been a very sophisticated field of
linguistics,  philosophy and logics, with many different approaches arguing for
and against each other. This section will not try to give an exhausting overview
of the field of question semantics since this would be a task of considerable
size and by far beyond the scope of this thesis\footnote{For a good summary of
some of the literature, particularly the issue of whether questions are a
concern of pragmatics or semantics, consider \cite[section 2 and 3]{gs:q}.}.
Instead, it was chosen to roughly follow the work by Groenendijk and Stokhof,
who present one of the best-established and recognized theories in the field.
In parts, the framework of the next sections is lending from other places, where
deemed appropriate or practical. As such, the theoretical framework of \acurt{}
is not going to be an exact realization of any particular approach to question
semantics, but rather an accumulation of interesting, and, most of all,
practicable concepts from different sources. The main sources, however, are
\cite{gs:q} and \cite{gs:sqpa}.

@@

\begin{termin}
  An \term{interrogative sentence}, or simply \term{interrogative}, is a
  type of sentence in natural language
  distinguished by certain characteristics
  \footnote{To quote \cite{gs:q}: \quotesf{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.},
  whereas an \term{interrogative act} refers to the utterance of an
  \term{interrogative}. Finally, a
  \term{question} will denote the semantic entity uttered by the
  \term{interrogative act}, and is introduced in section
  \ref{sec:protoq}
\end{termin}

@@

Example \ref{ex:yesno} is considered an \term{alternative question} or
\term{yes/no-interrogative} since only \emph{yes} and \emph{no} are possible
answers to this interrogative. \ref{ex:const}, on the other hand, is considered a
\term{\wh-question} or \term{constituent interrogative}, since one of the
sentences' nominal constituents is replaced by a \wh-pronoun, prompting an
enumeration of possible fillers for the resulting gap as an answer.

The following abbreviations are used: 
\abbr{Ty2}, for two-sorted type logic\footnote{See \cite{gallin:ty2} for a
discussion and \cite{gs:sqpa} for its usage in describing the semantics of
questions}, \abbr{IL}, for intensional logic\footnote{As used by
\cite{ptq}. See also \cite{gamut:2} for an excellent introduction to \abbr{IL}.},
\abbr{FOL}  is used as an abbreviation for first order logic. As an alternative
term for \term{possible world}, sometimes the expression \term{index} is used.

@@

Some research on the logic of questions (sometimes also referred to as eroteric
logic) suggests that the treatment of interrogatives demands its own formalism.
\acurt{} employs a formalism that departs from both \abbr{IL} and \abbr{Ty2} in
some ways. Most notably, the notion of a \term{model} is changed as follows.

@@

Burhans and Shapiro
